---
title: "CFPB Complaint Analysis"
author: "Jennifer Cooper"
date: "3/19/2020"
output:
  pdf_document: default
  html_document:
    code_folding: none
    df_print: paged
    highlight: tango
    theme: flatly
    toc: yes
    toc_depth: 2
params:
  interactive: FALSE
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    out.width = "100%",
    fig.align = "center")
```

```{r}
library("tidyverse")
library("readxl")
# install.packages("DataExplorer")
library("DataExplorer")
library("tidyquant")
library("ggrepel")
```

```{r}
# Copy and paste file location below, between quotation marks. Note double back slash(es). Note:  This is a big file, so it will take some time to load.

# Source: https://catalog.data.gov/dataset/consumer-complaint-database#topic=consumer_navigation

# Additional info:  https://www.consumerfinance.gov/data-research/consumer-complaints/

cfpb <- read.csv("C:\\Users\\jenni\\Desktop\\Job Search 2020\\Interview Prep\\Baylor\\complaints.csv", stringsAsFactors = FALSE)
# Note:  Data last updated -- January 27, 2020

# Data Dictionary: https://cfpb.github.io/api/ccdb/fields.html
```


## Business Problem

Healthcare is increasingly putting the patient/ consumer in focus. While there are existing ways/ sources providing a means to file / store / access a complaint against a provider, the purpose of this analysis is to examine consumer complaints regarding financial products and services to generate a summary / insight report that a company in the Dallas-Fort Worth area could use to glean value (e.g., get a general idea of the number and type of complaints consumers may file, as well as the channels they use, and how companies respond), based on Consumer Financial Protection Bureau (CFPB) Consumer Complaint data.

__Assumption(s)__:  Results could be used to benchmark the number / type of complaints made to healthcare providers and potentially improve responses to patient / consumer complaints in healthcare.

__Objective(s) & Scope__:

a. See if there is any data about consumer complaints by industry to gain context:

   https://consumerfed.org/press_release/nations-top-consumer-complaints-2/
  
   https://consumerfed.org/wp-content/uploads/2019/07/Top-Consumer-Complaints-Report.pdf
  
b. Examine CFPB data:

   1.) Summarize complaints by product
   2.) Review the types of complaints (e.g., issues)
   3.) Take a look at complaints by channel and state to detect any interesting patterns
   4.) Assess company response
   
c. Scope:

   1.) Time investment: ~ <= 8 hours (Will keep things high level)
   2.) Deliverable:  Code, Report (PDF, HTML, etc.); any instructions, supplemental data, needed to reproduce results.
  

## Summary of Findings

Consumers have sent over 1.5M complaints to the CFPB since March, 2017. This is over 45K complaints / month. 

The finacial products they complain most about are: credit bureau related services, mortgages and debt collection practices, as well as credit cards. 

More often than not, they submit their complaints online, are likely to be from California, Florida, Texas, New York, and Georgia and take issue with incorrect information, mortgage loan modifications and collection practices, as well as foreclosure matters. 

After a complaint is submitted by a consumer, the company will usually respond in 15 days, resulting in a "Closed with explanation" response without any monetary relief to the consumer.

Although there is a lot of debate and controversy surrounding the CFPB, in one consumer's view the process of "filing a complaint and getting a response from a financial institution [is still] very effective". (Source: [HTML Reports](https://thepointsguy.com/news/filing-consumer-financial-protection-bureau-complaint/)    

Data from the Consumer Federation of America suggests auto salestop complaints for across all industries, and "Health Products / Services" come in sixth (6th), with "misleading claims", "unlicensed practioners", "failure to deliver", and "medical billing issues" topping the list of complaints. 


## Recommendations & Next Steps

Before drawing any conclusions, it's recommended that companies reading this material consider their own industry, products and processes. Further, benchmarking complaints in said industry would be helpful. 

In the absence of this information, proactively surveying customers and / or consumer market research panels from survey providers like Zoomerang might be helpful to understand attitudes, relative to specific products and services.

## Data Analysis

Below is the supporting analysis conducted on the CFPB data. To reveal code, click on "Code" in the upper right hand corner of the code "chunk". To "Hide", click on "Hide". Note:  To show or hide all data, you can also click on the "Code" drop-down in the upper right hand corner of this document.

### Data preparation

The file used for this analysis comes from the data.gov website:

[HTML Reports](https://catalog.data.gov/dataset/consumer-complaint-database#topic=consumer_navigation)

The data spans complaints received from the Consumer Financial Protection Bureau between 3/23/2017 and 1/27/20.  There are 19 variables and ~ 1.5M observations in the data. To make the code run more efficiently and for the sake of time, I have taken a random sample of 15k records from the dataset.

```{r}
# Take a peek of all fields in the data & record number of variables and observations

#    Observations:   There are 19 variables or columns   
#         Records:   There are 1,531,308 obesrvations or complaints

glimpse(cfpb)
```

```{r}
# Since there's a large amount of data, I'm going to take a sample.
# set.seed() ensures reproducibility
# I am going to randomly sample 15,000 observations out of ~ 1.5M records

set.seed(12345)
cfpb_2 <- sample_n(cfpb,15000)
```

```{r, fig.height=8, fig.width=12}
# Get to know the data (look for any missing values and plot stats, as well as product complaint frequency)

# Note:   No missing values 

introduce(cfpb_2)

# Note:   Most columns are discrete

plot_intro(cfpb_2)
```

```{r}
# Rename columns and isolate fields needed; convert some to factors to make for easier graphing

cfpb_sample <- cfpb_2 %>%
    rename(date_received                   = Date.received,
           financial_product               = Product,
           financial_subproduct            = Sub.product,
           consumer_issue                  = Issue,
           consumer_subissue               = Sub.issue,
           consumer_complaint_comments     = Consumer.complaint.narrative,
           company_public_response         = Company.public.response,
           company                         = Company,
           consumer_state                  = State,
           consumer_zip                    = ZIP.code,
           complaint_search_tag            = Tags,
           consumer_publication_consent    = Consumer.consent.provided.,
           complaint_submission_channel    = Submitted.via,
           date_sent_to_company            = Date.sent.to.company,
           company_complaint_response      = Company.response.to.consumer,
           company_timely_response_flag    = Timely.response.,
           consumer_response_disputed      = Consumer.disputed.,
           complaint_id                    = Complaint.ID) %>%
    
# In the interest of time, going to focus on high level categories, not "sub" categories
  
    select(complaint_id,
           financial_product, 
           consumer_issue, 
           company_public_response,
           company_complaint_response,
           consumer_state,
           complaint_submission_channel
    )  

# Check out new dataframe; verify 7 fields isolated above

glimpse(cfpb_sample)
```

### Explore Data

__Complaints by Financial Product (Focus on top 10)__

"Credit Reporting", "Credit Repair Services", or "Other Personal Consumer Reports", followed by "Mortgage", and "Debt Collection" receive the most complaints.

```{r, fig.height=8, fig.width=12}
cfpb_sample %>%
    select(financial_product, complaint_id) %>%
    group_by(financial_product) %>%
    summarise(total_complaints = n()) %>%
    ungroup() %>%
    arrange(-total_complaints) %>%
    head(10) %>%
    mutate(financial_product = financial_product %>% as_factor() %>% fct_reorder(total_complaints)) %>%
    
    ggplot(aes(financial_product, total_complaints)) +
    geom_col(fill = "#2c3e50") +
    ggrepel::geom_label_repel(aes(label = total_complaints), size = 5) +
    theme_tq() +
    labs(title    = "Top 10 Financial Products by Number of Complaints",
         subtitle = "The financial product receiving the most complaints is 'Credit Bureau Services'",
         x        = "Total Complaints",
         y        = "Financial Product") +
    coord_flip()
```

__Complaints by Consumer Issue (Focus on top 10)__  

The top issues consumers have with financial products range from "Incorrect information on your report" to "Communication tactics".

```{r, fig.height=8, fig.width=12}
cfpb_sample %>%
    select(consumer_issue, complaint_id) %>%
    group_by(consumer_issue) %>%
    summarise(total_complaints = n()) %>%
    ungroup() %>%
    arrange(-total_complaints) %>%
    head(10) %>%
    mutate(consumer_issue = consumer_issue %>% as_factor() %>% fct_reorder(total_complaints)) %>%
    
    ggplot(aes(consumer_issue, total_complaints)) +
    geom_col(fill = "#2c3e50") +
    ggrepel::geom_label_repel(aes(label = total_complaints), size = 5) +
    theme_tq() +
    labs(title    = "Top 10 Consumer Issues by Number of Complaints",
         subtitle = "The biggest issue consumers have is 'Incorrect information on their report'",
         x        = "Total Complaints",
         y        = "Consumer Issue") +
    coord_flip()
```
 
__Consumer Preferences re: Channel to Submit Complaint__ 

Consumers prefer to submit their complaints throug the web.

```{r, fig.height=8, fig.width=12}
cfpb_sample %>%
    select(complaint_submission_channel, complaint_id) %>%
    group_by(complaint_submission_channel) %>%
    summarise(total_complaints = n()) %>%
    ungroup() %>%
    arrange(-total_complaints) %>%
    mutate(complaint_submission_channel = complaint_submission_channel %>% as_factor() %>% fct_reorder(total_complaints)) %>%
    
    ggplot(aes(complaint_submission_channel, total_complaints)) +
    geom_col(fill = "#2c3e50") +
    ggrepel::geom_label_repel(aes(label = total_complaints), size = 5) +
    theme_tq() +
    labs(title    = "Consumer Complaints by Channel",
         subtitle = "Most complaints are submitted through the Web",
         x        = "Total Complaints",
         y        = "Channel") +
    coord_flip()
```

__Top Company Responses (Focus on Top 10)__ 

The most prevalent responses to consumer complaints reported by companies is "Closed with explanation" and "Closed with non-monetary relief". 

```{r, fig.height=8, fig.width=12}
cfpb_sample %>%
    select(company_complaint_response, complaint_id) %>%
    group_by(company_complaint_response) %>%
    summarise(total_complaints = n()) %>%
    ungroup() %>%
    arrange(-total_complaints) %>%
    head(10) %>%
    mutate(company_complaint_response = company_complaint_response %>% as_factor() %>% fct_reorder(total_complaints)) %>%
    
    ggplot(aes(company_complaint_response, total_complaints)) +
    geom_col(fill = "#2c3e50") +
    ggrepel::geom_label_repel(aes(label = total_complaints), size = 5) +
    theme_tq() +
    labs(title    = "Top 10 Company Responses by Number of Complaints",
         subtitle = "The most popular company response is 'Closed with explanation', followed by 'Closed with Non-monetary Relief' and 'Closed with Monetary Relief'",
         x        = "Total Complaints",
         y        = "Company Response") +
    coord_flip()
```

__Top States for Complaints (Focus on Top 10)__

Most complaints are coming from California, Florida, Texas, New York, and Georgia.

```{r, fig.height=8, fig.width=12}
cfpb_sample %>%
    select(consumer_state, complaint_id) %>%
    group_by(consumer_state) %>%
    summarise(total_complaints = n()) %>%
    ungroup() %>%
    arrange(-total_complaints) %>%
    head(10) %>%
    mutate(consumer_state = consumer_state %>% as_factor() %>% fct_reorder(total_complaints)) %>%
    
    ggplot(aes(consumer_state, total_complaints)) +
    geom_col(fill = "#2c3e50") +
    ggrepel::geom_label_repel(aes(label = total_complaints), size = 5) +
    theme_tq() +
    labs(title    = "Top 10 States by Number of Complaints",
         subtitle = "Most complaints are coming from California, followed by Florida, Texas, New York and Georgia",
         x        = "Total Complaints",
         y        = "Company Response") +
    coord_flip()
```

